---
title: "diseasystore: Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diseasystore: Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To showcase the performance of `diseasystore` on different database backends, we include this vignette that summarises
a simple benchmark:
A sample dataset is created based on the `datasets::mtcars` dataset. This data is repeated 10 times and given a unique
ID (the row number of the data).

A simple `diseasystore` is built around this data, with two `FeatureHandlers`, one each for the `cyl` and `vs`
variables.
```{r, eval = FALSE}
DiseasystoreMtcars <- R6::R6Class(
  classname = "DiseasystoreBase",
  inherit = DiseasystoreBase,
  private = list(
    .ds_map = list("n_cyl" = "mtcars_cyl", "vs" = "mtcars_vs"),
    mtcars_cyl = FeatureHandler$new(
      compute = function(start_date, end_date, slice_ts, source_conn) {
        out <- benchmark_data |>
          dplyr::transmute(
            "key_car" = .data$car, "n_cyl" = .data$cyl,
            "valid_from" = Sys.Date() - lubridate::days(2 * .data$r - 1),
            "valid_until" = .data$valid_from + lubridate::days(2)
          )
        return(out)
      },
      key_join = key_join_sum
    ),
    mtcars_vs = FeatureHandler$new(
      compute = function(start_date, end_date, slice_ts, source_conn) {
        out <- benchmark_data |>
          dplyr::transmute(
            "key_car" = .data$car, .data$vs,
            "valid_from" = Sys.Date() - lubridate::days(2 * .data$r),
            "valid_until" = .data$valid_from + lubridate::days(2)
          )
        return(out)
      },
      key_join = key_join_sum
    )
  )
)
```

Two separate benchmark functions are created.
The first benchmarking function tests the computation time of `$get_feature()` by computing first the `n_cyl` feature
then computing the `vs` feature, before finally deleting the computations from the database.

The second benchmarking function tests the computation time of `$key_join_features()` by joining the `vs` feature to the
`n_cyl` observation. Note that the `n_cyl` and `vs` are re-computed before the benchmarks are started and are not
deleted by the benchmarking function as was the case for the benchmark of `$get_feature()`.

The performance of these benchmark functions are timed with the `microbenchmark` package using 10 replicates.
All benchmarks are run on the same machine.

The results of the benchmark are shown graphically below (mean and standard deviation), where we compare the current
development version of `diseasystore` with the current CRAN version. In addition, we run the benchmarks with the
current development version of `SCDB` versus the current CRAN version since `SCDB` is used internally in `diseasystore`.

```{r, echo = FALSE, eval = rlang::is_installed("here")}
benchmark_location <- c(
  devtools::package_file("inst", "extdata", "benchmarks.rds"),
  system.file("extdata", "benchmarks.rds", package = "diseasystore"),
  here::here("inst", "extdata", "benchmarks.rds")
) |>
  purrr::discard(~ identical(., "")) |>
  purrr::pluck(1)

benchmarks <- readRDS(benchmark_location)

# Determine if the SHA is on main
sha <- benchmarks |>
  dplyr::distinct(version) |>
  dplyr::filter(!(version %in% c("CRAN", "main"))) |>
  dplyr::pull("version")

# Check local git history
on_main <- tryCatch({
  system(glue::glue("git branch main --contains {sha}"), intern = TRUE) |>
    stringr::str_detect(stringr::fixed("main")) |>
    isTRUE()
}, warning = function(w) {
  # If on GitHub, git is not installed and we assume TRUE.
  # This will render the vignette as it will look once merged onto main.
  return(identical(Sys.getenv("CI"), "true"))
})

# If the SHA has been merged, use as the "main" version and remove the other, older, main version
if (on_main) {
  benchmarks <- benchmarks |>
    dplyr::filter(version != "main") |>
    dplyr::mutate(version = dplyr::if_else(.data$version == "CRAN", "CRAN", "development"))
}
```

```{r Generate alt text for benchmarks, echo = FALSE, results = "hide"}
benchmark_alt_text <- benchmarks |>
  dplyr::filter(.data$version == "CRAN", .data$SCDB == "CRAN") |>
  dplyr::transmute(
    .data$benchmark_function,
    .data$database,
    "time_seconds" = .data$time / 1e9
  ) |>
  dplyr::mutate("backend" = stringr::str_remove(.data$database, r"{\sv[\w+\.]+$}")) |>
  dplyr::summarise(
    "mean_time_seconds" = round(mean(.data$time_seconds)),
    .by = c("benchmark_function", "backend")
  ) |>
  dplyr::summarise(
    "backends" = toString(.data$backend),
    .by = c("benchmark_function", "mean_time_seconds")
  ) |>
  dplyr::arrange(.data$benchmark_function, .data$mean_time_seconds) |>
  purrr::pmap_chr(
    \(benchmark_function, mean_time_seconds, backends) {
      glue::glue(
        "The {benchmark_function} benchmark takes {mean_time_seconds} seconds to run on the {backends} backends."
      )
    }
  ) |>
  purrr::reduce(paste, .init = "The results for the benchmarks of the CRAN versions are as follows:")
```

```{r, echo = FALSE, fig.alt = benchmark_alt_text, eval = rlang::is_installed("here")}

# Add note slow backends
slow_backends <- benchmarks |>
  dplyr::distinct(.data$database, .data$benchmark_function, .data$n) |>
  dplyr::filter(.data$n < max(.data$n), .by = c("database", "benchmark_function")) |>
  dplyr::select(!"n") |>
  dplyr::mutate("slow_backend" = TRUE)

benchmarks <- benchmarks |>
  dplyr::left_join(slow_backends, by = c("database", "benchmark_function")) |>
  dplyr::mutate("database" = paste0(.data$database, ifelse(isTRUE(.data$slow_backend), "*", ""))) |>
  dplyr::select(!"slow_backend")


# Mean and standard deviation (see ggplot2::mean_se())
mean_sd <- function(x) {
  mu <- mean(x)
  sd <- sd(x)
  data.frame(y = mu, ymin = mu - sd, ymax = mu + sd)
}

# Determine subgroups
groups <- setdiff(colnames(benchmarks), c("expr", "time", "benchmark_function", "database", "version", "n"))

# Apply "dodging" to sub-groups to show graphically
dodge <- ggplot2::position_dodge(width = 0.6)

# Insert newline into database name to improve rendering of figures
labeller <- ggplot2::as_labeller(\(l) stringr::str_replace_all(l, stringr::fixed(" v"), "\nv"))

g <- ggplot2::ggplot(
  benchmarks,
  ggplot2::aes(
    x = version,
    y = time / 1e9,
    group = !!switch(length(groups) > 0, as.symbol(groups)),
    color = !!switch(length(groups) > 0, as.symbol(groups))
  )
) +
  ggplot2::stat_summary(fun.data = mean_sd, geom = "pointrange", size = 0.5, linewidth = 1, position = dodge) +
  ggplot2::facet_grid(
    rows = ggplot2::vars(benchmark_function),
    cols = ggplot2::vars(database),
    scales = "free_y",
    labeller = labeller
  ) +
  ggplot2::labs(x = "Codebase version", y = "Time (s)") +
  ggplot2::theme(
    legend.position = "bottom",
    legend.justification = "left"
  )

if (nrow(slow_backends) > 1) {
  g <- g + ggplot2::labs(caption = "* IMPORTANT: Benchmark data halved for this backend!")
}

g
```
