---
title: "diseasystore: ECDC Respiratory Viruses Weekly"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diseasystore: ECDC Respiratory Viruses Weekly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasystore)
```

```{r hidden_options, include = FALSE}
if (rlang::is_installed("withr")) {
  withr::local_options("tibble.print_min" = 5)
  withr::local_options("diseasystore.verbose" = FALSE)
  withr::local_options("diseasystore.DiseasystoreEcdcRespiratoryViruses.pull" = FALSE)
} else {
  opts <- options(
    "tibble.print_min" = 5,
    "diseasystore.verbose" = FALSE,
    "diseasystore.DiseasystoreEcdcRespiratoryViruses.pull" = FALSE
  )
}

# We have a "hard" dependency for duckdb to render parts of this vignette
suggests_available <- rlang::is_installed("duckdb")
not_on_cran <- interactive() || identical(Sys.getenv("NOT_CRAN"), "true") || identical(Sys.getenv("CI"), "true")
```

The Respiratory viruses weekly repository contains weekly summaries of the
[erviss](https://www.ecdc.europa.eu/en/publications-data/european-respiratory-virus-surveillance-summary-erviss) data
for Influenza-Like Illness (ILI) and Acute Respiratory Infection (ARI).

This vignette shows how to use (some of) this data through the `diseasystore` package.

First, it is a good idea to clone the repository locally and store that location as an option for the package.

```{r download_data, eval = FALSE}
# First we set the path we want to use as an option
options(
  "diseasystore.DiseasystoreEcdcRespiratoryViruses.source_conn" =
    file.path("local", "path")
)

# Ensure folder exists
source_conn <- diseasyoption("source_conn", "DiseasystoreEcdcRespiratoryViruses")
if (!dir.exists(source_conn)) {
  dir.create(source_conn, recursive = TRUE, showWarnings = FALSE)
}

# Clone the repository
system2(
  "git",
  args = c(
    paste("-C", source_conn),
    "clone https://github.com/EU-ECDC/Respiratory_viruses_weekly_data"
  ),
  stdout = TRUE
)
```

```{r download_data_hidden, include = FALSE, eval = not_on_cran}
# In practice, it is best to make a local copy of the data which is stored in the "vignette_data" folder
# This folder can either be in the package folder (preferred, please create the folder) or in the tempdir()
local_conn <- purrr::detect("vignette_data", checkmate::test_directory_exists, .default = tempdir())

if (rlang::is_installed("withr")) {
  withr::local_options("diseasystore.DiseasystoreEcdcRespiratoryViruses.source_conn" = local_conn)
  withr::local_options("diseasystore.DiseasystoreEcdcRespiratoryViruses.n_max" = 1000)
} else {
  opts <- c(opts, options("diseasystore.DiseasystoreEcdcRespiratoryViruses.source_conn" = local_conn,
                          "diseasystore.DiseasystoreEcdcRespiratoryViruses.n_max" = 1000))
}

# Define the ECDC file to download
test_file <- "data/snapshots/2023-11-24_ILIARIRates.csv"

# Create folder and download file
dir.create(dirname(file.path(local_conn, test_file)), recursive = TRUE, showWarnings = FALSE)

if (!file.exists(file.path(local_conn, test_file))) {
  remote_file <- source_conn_github(
    diseasyoption("remote_conn", "DiseasystoreEcdcRespiratoryViruses"),
    test_file
  )

  readr::read_csv(remote_file, n_max = 1000, show_col_types = FALSE, progress = FALSE) |>
    readr::write_csv(file.path(local_conn, test_file))
}

# Check that the files are available after attempting to download
if (purrr::some(test_file, ~ !checkmate::test_file_exists(file.path(local_conn, .)))) {
  data_available <- FALSE
} else {
  data_available <- TRUE
}
```


The `diseasystores` require a database to store its features in.
These should be configured before use and can be stored in the packages options.
```{r configure_diseasystore, eval = FALSE}
# We define target_conn as a function that opens a DBIconnection to the DB
target_conn <- \() DBI::dbConnect(duckdb::duckdb())
options(
  "diseasystore.DiseasystoreEcdcRespiratoryViruses.target_conn" = target_conn
)
```


```{r configure_diseasystore_hidden, include = FALSE, eval = not_on_cran && suggests_available && data_available}
target_conn <- \() DBI::dbConnect(duckdb::duckdb())
if (rlang::is_installed("withr")) {
  withr::local_options("diseasystore.DiseasystoreEcdcRespiratoryViruses.target_conn" = target_conn)
} else {
  opts <- c(opts, options("diseasystore.DiseasystoreEcdcRespiratoryViruses.target_conn" = target_conn))
}
```


Once the files are downloaded and the target database is configured, we can initialize the `diseasystore` that uses the
ECDC Respiratory Viruses Weekly data.
```{r initializing_diseasystore, eval = not_on_cran && suggests_available && data_available}
ds <- DiseasystoreEcdcRespiratoryViruses$new()
```

```{r duckdb_suppress, include = FALSE, eval = not_on_cran && suggests_available && data_available}
# duckdb has decided to be very verbose about accessing tables, at least the first time you call a table,
# so we suppress the output here
dummy_table <- dplyr::copy_to(ds %.% target_conn, data.frame(a = 1, b = 2), name = SCDB::unique_table_name())
tbl <- purrr::quietly(dplyr::tbl)
t <- tbl(ds %.% target_conn, SCDB::id(dummy_table))
stopifnot("DuckDB no longer has verbose messages"={!identical(t$messages, character(0))}) # If this fails, duckdb has removed the message and we can remove this section
```

Once configured such, we can use the feature store directly to get data.
```{r using_diseasystore_example_1, eval = not_on_cran && suggests_available && data_available}
# We can see all the available features in the feature store
ds$available_features
```

If you have locally cloned the repository, you may consider pulling data from the repository manually and disabling
automatic pulls. Otherwise, the `diseasystore` will check for updates every time you request a feature.
```{r using_diseasystore_example_2, eval = FALSE}
# Manually update the repository
system2(
  "git",
  args = paste("-C", diseasyoption("source_conn", "DiseasystoreEcdcRespiratoryViruses")),
  stdout = TRUE
)

# Disable automatic pulls
options("diseasystore.DiseasystoreEcdcRespiratoryViruses.pull" = FALSE)
```


```{r using_diseasystore_example_3, eval = not_on_cran && suggests_available && data_available}
# And then retrieve a feature from the feature store
ds$get_feature(feature = "iliari_rates",
               start_date = as.Date("2023-01-01"),
               end_date = as.Date("2023-03-01"))
```


```{r cleanup, include = FALSE}
if (exists("ds")) rm(ds)
gc()
if (!rlang::is_installed("withr")) {
  options(opts)
}
```
