---
title: "diseasystore: Google Health COVID-19 Open Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diseasystore: Google Health COVID-19 Open Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasystore)
```

```{r hidden_options, include = FALSE}
if (rlang::is_installed("withr")) {
  withr::local_options("tibble.print_min" = 5)
  withr::local_options("diseasystore.verbose" = FALSE)
} else {
  opts <- options("tibble.print_min" = 5, "diseasystore.verbose" = FALSE)
}

# We have a "hard" dependency for duckdb to render parts of this vignette
suggests_available <- rlang::is_installed("duckdb")
not_on_cran <- interactive() || identical(Sys.getenv("NOT_CRAN"), "true") || identical(Sys.getenv("CI"), "true")
```

The Google COVID-19 data repository is a comprehensive open repository of COVID-19 data.

This vignette shows how to use (some of) this data through the `diseasystore` package.

First, it is a good idea to copy the relevant Google COVID-19 data files locally and store that location as an option
for the package.
`DiseasystoreGoogleCovid19` uses only the age-stratified metrics for COVID-19, so only a subset of the repository is
needed to download.

```{r download_data, eval = FALSE}
# First we set the path we want to use as an option
options(
  "diseasystore.DiseasystoreGoogleCovid19.source_conn" =
    file.path("local", "path")
)

# Ensure folder exists
source_conn <- diseasyoption("source_conn", "DiseasystoreGoogleCovid19")
if (!dir.exists(source_conn)) {
  dir.create(source_conn, recursive = TRUE, showWarnings = FALSE)
}

# Define the Google files to download
google_files <- c("by-age.csv", "demographics.csv", "index.csv", "weather.csv")

# Download each file and compress them to reduce storage
purrr::walk(google_files, ~ {
  url <- paste0(diseasyoption("remote_conn", "DiseasystoreGoogleCovid19"), .)

  destfile <- file.path(
    diseasyoption("source_conn", "DiseasystoreGoogleCovid19"),
    .
  )

  if (!file.exists(destfile)) {
    download.file(url, destfile)
  }
})
```

```{r download_data_hidden, include = FALSE, eval = not_on_cran}
# The files we need are stored remotely in Google's API
google_files <- c("by-age.csv", "demographics.csv", "index.csv", "weather.csv")
remote_conn <- diseasyoption("remote_conn", "DiseasystoreGoogleCovid19")

# In practice, it is best to make a local copy of the data which is stored in the "vignette_data" folder
# This folder can either be in the package folder (preferred, please create the folder) or in the tempdir()
local_conn <- purrr::detect("vignette_data", checkmate::test_directory_exists, .default = tempdir())

if (rlang::is_installed("withr")) {
  withr::local_options("diseasystore.DiseasystoreGoogleCovid19.source_conn" = local_conn)
  withr::local_options("diseasystore.DiseasystoreGoogleCovid19.n_max" = 1000)
} else {
  opts <- c(opts, options("diseasystore.DiseasystoreGoogleCovid19.source_conn" = local_conn,
                          "diseasystore.DiseasystoreGoogleCovid19.n_max" = 1000))
}

# Then we download the first n rows of each data set of interest
try({
  purrr::discard(google_files, ~ checkmate::test_file_exists(file.path(local_conn, .))) |>
    purrr::walk(\(file) {
      paste0(remote_conn, file) |>
        readr::read_csv(n_max = 1000, show_col_types = FALSE, progress = FALSE) |>
        readr::write_csv(file.path(local_conn, file))
    })
})

# Check that the files are available after attempting to download
if (purrr::some(google_files, ~ !checkmate::test_file_exists(file.path(local_conn, .)))) {
  data_available <- FALSE
} else {
  data_available <- TRUE
}
```


The `diseasystores` require a database to store its features in.
These should be configured before use and can be stored in the packages options.
```{r configure_diseasystore, eval = FALSE}
# We define target_conn as a function that opens a DBIconnection to the DB
target_conn <- \() DBI::dbConnect(duckdb::duckdb())
options(
  "diseasystore.DiseasystoreGoogleCovid19.target_conn" = target_conn
)
```

```{r configure_diseasystore_hidden, include = FALSE, eval = not_on_cran && suggests_available && data_available}
target_conn <- \() DBI::dbConnect(duckdb::duckdb())
if (rlang::is_installed("withr")) {
  withr::local_options("diseasystore.DiseasystoreGoogleCovid19.target_conn" = target_conn)
} else {
  opts <- c(opts, options("diseasystore.DiseasystoreGoogleCovid19.target_conn" = target_conn))
}
```


Once the files are downloaded and the target DB is configured, we can initialize the `diseasystore` that uses the
Google COVID-19 data.
```{r initializing_diseasystore, eval = not_on_cran && suggests_available && data_available}
ds <- DiseasystoreGoogleCovid19$new()
```

Once configured such, we can use the feature store directly to get data.
```{r using_diseasystore_example_1, eval = not_on_cran && suggests_available && data_available}
# We can see all the available features in the feature store
ds$available_features
```

```{r using_diseasystore_example_2, eval = not_on_cran && suggests_available && data_available}
# And then retrieve a feature from the feature store
ds$get_feature(feature = "n_hospital",
               start_date = as.Date("2020-01-01"),
               end_date = as.Date("2020-06-01"))
```


```{r cleanup, include = FALSE}
if (exists("ds")) rm(ds)
gc()
if (!rlang::is_installed("withr")) {
  options(opts)
}
```
