---
title: "Extending diseasystore - simulist example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending diseasystore - simulist example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasystore)
```

```{r hidden_options, include = FALSE}
if (rlang::is_installed("withr")) {
  withr::local_options("tibble.print_min" = 5)
  withr::local_options("diseasystore.verbose" = FALSE)
} else {
  opts <- options("tibble.print_min" = 5, "diseasystore.verbose" = FALSE)
}

# We have a "hard" dependency for DuckDB to render parts of this vignette
suggests_available <- rlang::is_installed("duckdb")
```

# Introduction
If you haven't already, please consult the `vignette("extending-diseasystore")` which explains the concepts of the
`diseasystore`s in detail.

In this example, we go through how to create a `diseasystore` that implements individual level data.
For this purpose, we use a synthetic data set that we have generated using the `{simulist}` package.
This data set is a synthetic line list from a disease outbreak.

```{r simulist_data}
simulist_data
```

Given such a data set, the aim is to create a `diseasystore` that can be used to retrieve features from this data set.


# Feature identification

To do this, we need to identify the "features" that we want to store in the `diseasystore`.
In the data, there are some demographic features such as "sex" and "age" and there are disease specific features such
as "date_onset", "date_admission", et cetera.

In this example we will implement the following features:

* Demographic features
  * `birth` - The birth date of the individuals
  * `age` - The (time-dependent) age of the individuals
  * `sex` - The sex of the individuals

* Disease features
  * `n_positive` - The positive tests performed on the individuals
  * `n_admission` - The time of admission among the individuals
  * `n_hospital` - The time of hospitalisation among the individuals

Notice that the naming of the features differ by the prefix `n_` versus no prefix.
This dichotomy separates "stratifications" from "observables". Either is considered a "feature" in the diseasystore,
but the type of data (stratification vs. observable) is used when we join the data in `$key_join_features()`.
However, there are no differences in the implementation of the features --- only in the naming.


With the features identified, we begin by assigning a `FeatureHandler` to each feature, and design the map between
the feature names and the responsible `FeatureHandlers`:

```{r ds_map, eval = FALSE}
DiseasystoreSimulist <- R6::R6Class(
  classname = "DiseasystoreSimulist",
  inherit = DiseasystoreBase,

  ...

  private = list(
    .ds_map = list(
      "birth"       = "simulist_birth",
      "sex"         = "simulist_sex",
      "age"         = "simulist_age",
      "n_positive"  = "simulist_positive",
      "n_admission" = "simulist_admission",
      "n_hospital"  = "simulist_hospital"
    ),
    .label = "Simulist Synthetic Data",

  ...
)
```
Notice that the names for the `FeatureHandlers` have the prefix "simulist_". The features are stored in the data base
using these names, so we include the prefix to avoid any potential conflicts with other features in the database.
For example, the hospitalisation in `DiseasystoreGoogleCovid19` are stored in a table called "google_covid_19_hospital".
Had both `diseasystores` omitted the prefixes, they would both attempt to write to the same "hospital" table in the
database.

# Feature implementation
With the features identified, it is time to implement them in the `diseasystore`.

That is, we need to construct the `FeatureHandlers` for each feature.

## The `birth` `FeatureHandler`
We start by implementing the `birth` feature.

For this example, we will show first the code that implements the `FeatureHandler` and then describe the code in detail.
```{r feature_handler_birth, eval = FALSE}
private = list(
  ...

  # The "birth" feature contains the birth dates of the individuals and is used later
  # to compute the age of the individuals at any given time.
  simulist_birth = FeatureHandler$new(
    compute = function(start_date, end_date, slice_ts, source_conn, ...) {

      out <- simulist_data |>
        dplyr::transmute(
          "key_pnr" = .data$id,
          "birth" = .data$birth,
          "valid_from" = .data$birth,
          "valid_until" = .data$date_death + lubridate::days(1)
        ) |>
        dplyr::filter(
          {{ start_date }} < .data$valid_until,
          .data$valid_from <= {{ end_date }}
        )

      return(out)
    },
    key_join = key_join_count
  ),

  ...
)
```

Notice that we create a new instance of `FeatureHandler` and assign it to the `simulist_birth` field in the private
list.

Lets focus on the `FeatureHandler` we create:
Here, we need to define how the feature is computed (that is, we implement the `compute()` function).

The signature of this function is fixed, and must take the arguments `start_date`, `end_date`, `slice_ts`,
`source_conn`, but can then take additional arguments in `...`.

The `birth` feature is a simple feature that we can extract directly from the underlying data set: `simulist_data`.

In addition to the feature values themselves (the values stored in the `birth` column) we need to include keying data
that tells `diseasystore` what other features this feature can be joined to.
Since we are working with individual level data, we can use a personal identification number, "pnr", as the key.
The simulist data already contains a unique ID for each individual, so we use this as the personal identification
number.

In addition, we need to specify the time period that the feature covers.
In this case, a birth date is assigned at birth, so we use this as the `valid_from` date.
For the expiration date of the information, we could either use `NA`, since the birth date is always valid, or
we could use the date of death, if the individual has died.
In this implementation we use the latter, since this allows for an easier implementation of the age of the individuals
later on.

Notice that `valid_until` is set as the date after death. The validity period of the features in `diseasystore` is
the interval $t \in \left[\textrm{valid_from}, \textrm{valid_until}\right)$.

After computing the feature, we filter to ensure that the data is returned only for the requested time period.

Finally, we also how the feature can be summarised. That is, we need to specify the type of `key_join` the feature
should use.

To illustrate how to determine the type of `key_join`, we imagine that we want to determine the number of people who
are born on each date. Given our data set, we could perform the following operations to summarize the data:

```{r key_join_birth}
simulist_data |>
  dplyr::count(lubridate::year(.data$birth))
```

That is, our feature has a record for each individual, and to summarise the feature, we need to `count` the individuals
within each group. This is different from semi-aggregated data where we would `sum` the values within each group.


## The `sex` `FeatureHandler`
We continue by implementing the `sex` feature.

Like the `birth` feature, this is a simple feature that we can mostly extract directly from the data set:

```{r feature_handler_sex, eval = FALSE}
private = list(
  ...

  # The "sex" feature simply stores the sex from the simulist data
  simulist_sex = FeatureHandler$new(
    compute = function(start_date, end_date, slice_ts, source_conn, ds, ...) {

      out <- simulist_data |>
        dplyr::right_join( # Join with birth data to validity period
          ds$get_feature("birth", start_date, end_date, slice_ts),
          by = c("id" = "key_pnr"),
          copy = TRUE
        ) |>
        dplyr::transmute(
          "key_pnr" = .data$id,
          "sex" = dplyr::if_else(.data$sex == "m", "Male", "Female"),
          .data$valid_from, .data$valid_until # Use values from birth feature
        )

      # No need to filter to ensure the data is only for the requested time period.
      # Since we right join with the birth feature, the validity period is already filtered.

      return(out)
    },
    key_join = key_join_count
  ),

  ...
)
```

Notice that we again create a new instance of `FeatureHandler` and assign it to the `simulist_sex` field in the
private list.

The computation of the feature is mostly straight forward, but we would like to use the validity period computed
in the `birth` feature (the same logic applies for the validity period of the "sex" information as the "birth"
information)

Therefore, we immediately see a difference in the function signature of the `compute()` function, where we no include
the `ds` argument which, like the others, are passed to the feature handler by `diseasystore` when we compute features.
This argument holds a reference to the `diseasystore` we are building, so we can use it to retrieve other features
(in this case, the `birth` feature).

Once we have the `birth` feature, we combine it with the `simulist_data` and extract a human-readable label for the
`sex`.


## The `age` `FeatureHandler`
The `age` feature is a bit more complex than the previous features, as it is a time-dependent feature.

Lets first look at the code:

```{r feature_handler_age, eval = FALSE}
private = list(
  ...

  # The "age" feature computes the age of the individuals throughout the study period
  simulist_age = FeatureHandler$new(
    compute = function(start_date, end_date, slice_ts, source_conn, ds, ...) {

      # Using birth date, compute the age at the start of the study period
      age <- ds$get_feature("birth", start_date, end_date, slice_ts) |>
        dplyr::mutate(
          age_at_start = as.integer(
            !!age_on_date("birth", start_date, conn = ds %.% target_conn)
          )
        ) |>
        dplyr::compute()

      # Now, compute the next birth dates of the individual
      # (as many as we need to cover the study period)
      # and compute the age of the individuals throughout the study period with their
      # birth dates denoting the starts and ends of the validity periods.
      out <- purrr::map(
        seq.int(
          from = 0,
          to = ceiling(lubridate::interval(start_date, end_date) / lubridate::years(1))
        ),
        ~ age |>
          dplyr::mutate(
            # The age for this iteration of the age computation loop
            "age" = .data$age_at_start + .x
          ) |>
          dplyr::mutate( # Split to make the "age" column available for the next mutate
            # Compute the birthday for the age
            "birthday" = !!add_years("birth", "age", conn = ds %.% target_conn)
          ) |>
          dplyr::mutate( # Again, split to make "birthday" available for the next mutate
            # And when that age is not valid
            "next_birthday" = !!add_years("birthday", 1, conn = ds %.% target_conn)
          ) |>
          dplyr::filter( # Remove the birthdays that fall outside of the study period
            .data$birthday <= {{ end_date }},
            .data$birthday < .data$valid_until | is.na(.data$valid_until)
          ) |>
          dplyr::transmute( # We assign the birth dates as the validity periods
            "key_pnr" = .data$key_pnr,
            "age" = .data$age,
            "valid_from" = .data$birthday,
            "valid_until" = pmin(
              .data$valid_until,
              .data$next_birthday,
              na.rm = TRUE
            )
          )
      ) |>
        purrr::reduce(dplyr::union_all) # Collapse to a single dataset

      return(out)
    },
    key_join = key_join_count
  ),

  ...
)
```

As with the `sex` feature, this feature also depends on the `birth` feature, so we include the `ds` argument in the
function signature of `compute()`.

Since the age is time-dependent, we need to compute the age of the individuals at the start of the study period (i.e.
the period requested by the user), and compute their age throughout the study period.

The age at the start of the study period is computed directly by from the birth date of the individuals using the
age helper function `age_on_date()`[^1].

Once the age at start is known, we determine the length of the study period (in years) and compute the birth days of
the individuals throughout the period[^2]. Each birth day has an associated age and a validity period which runs from
the birth day to the next birth day. Notice that we sometimes compute birth days that fall outside of the study period,
which we filter out before combining to a single data set with `dplyr::union_all`.

In all, we end with a data set containing the time-dependent age of the individuals throughout the study period.

## The `n_positive` `FeatureHandler`
The `n_positive` feature is a simple feature that we can extract directly from the data set with a little filtering:

```{r feature_handler_n_positive, eval = FALSE}
private = list(
  ...

  # The "n_positive" feature contains the positive tests taken by the individuals
  simulist_positive = FeatureHandler$new(
    compute = function(start_date, end_date, slice_ts, source_conn, ...) {

      out <- simulist_data |>
        dplyr::filter(.data$case_type == "confirmed") |>
        dplyr::transmute(
          "key_pnr" = .data$id,
          "valid_from" = .data$date_onset,
          "valid_until" = .data$valid_from + lubridate::days(1)
        ) |>
        dplyr::filter(
          {{ start_date }} < .data$valid_until,
          .data$valid_from <= {{ end_date }}
        )

      return(out)
    },
    key_join = key_join_count
  ),

  ...
)
```

## The `n_hospital` `FeatureHandler`
Like the `n_positive` feature, the `n_hospital` is straightforward to implement:
```{r feature_handler_n_hospital, eval = FALSE}
private = list(
  ...

  # The "n_hospital" feature contains the hospitalizations of the individuals
  simulist_hospital = FeatureHandler$new(
    compute = function(start_date, end_date, slice_ts, source_conn, ds, ...) {

      out <- simulist_data |>
        dplyr::filter(
          .data$case_type == "confirmed",
          !is.na(.data$date_admission)
        ) |>
        dplyr::transmute(
          "key_pnr" = .data$id,
          "valid_from" = .data$date_admission,
          "valid_until" = .data$date_discharge + lubridate::days(1)
        ) |>
        dplyr::filter(
          {{ start_date }} < .data$valid_until,
          .data$valid_from <= {{ end_date }}
        )

      return(out)
    },
    key_join = key_join_count
  ),

  ...
)
```

## The `n_admission` `FeatureHandler`
For the `n_admission` feature, we use the already computed `n_hospital` feature to compute the admissions.
We could have computed directly from the simulist data, but since this features is really just a transformation of
the `n_hospital` feature, we use the `n_hospital` feature to compute the admissions.

Since we modify the `valid_until` field, we need to filter again to ensure that the data is only for the requested
time period.

```{r feature_handler_n_admission, eval = FALSE}
private = list(
  ...

  # The "n_admission" feature contains the admissions of the individuals
  # We here use the "n_hospital" feature to compute the admissions since the admission
  # is an entry for the first date of hospitalisation
  simulist_admission = FeatureHandler$new(
    compute = function(start_date, end_date, slice_ts, source_conn, ds, ...) {

      out <- ds$get_feature("n_hospital", start_date, end_date, slice_ts) |>
        dplyr::mutate("valid_until" = .data$valid_from + 1L) |>
        dplyr::filter({{ start_date }} < .data$valid_until) # valid_from filtered in n_hospital

      return(out)
    },
    key_join = key_join_count
  )

  ...
)
```


# The final product
We are finally ready to see the `diseasystore` in action.

## Configuring the `diseasystore`
All `diseasystores` require a database to store its features in.
For convenience, we here set the `target_conn` option for all `diseasystores`[^3].
```{r configure_diseasystore, eval = FALSE}
# We define target_conn as a function that opens a DBIconnection to the DB
target_conn <- \() DBI::dbConnect(duckdb::duckdb())
options(
  "diseasystore.target_conn" = target_conn
)
```

```{r configure_diseasystore_hidden, include = FALSE, eval = suggests_available}
target_conn <- \() DBI::dbConnect(duckdb::duckdb())
if (rlang::is_installed("withr")) {
  withr::local_options("diseasystore.DiseasystoreSimulist.target_conn" = target_conn)
} else {
  opts <- c(opts, options("diseasystore.DiseasystoreSimulist.target_conn" = target_conn))
}
```

With the `target_conn` option set, we can easily create a new instance of the `diseasystore`[^4].
```{r initializing_diseasystore, eval = suggests_available}
ds <- diseasystore:::DiseasystoreSimulist$new()
```

## Retrieving features
We can see that the features we implemented are available:
```{r ds_available_features, eval = suggests_available}
ds$available_features
```

And we can retrieve the individual features
```{r get_feature_sex, eval = suggests_available}
ds$get_feature(
  feature = "sex",
  start_date = ds$min_start_date,
  end_date = ds$max_end_date
)
```

```{r get_feature_n_positive, eval = suggests_available}
ds$get_feature(
  feature = "n_positive",
  start_date = ds$min_start_date,
  end_date = ds$max_end_date
)
```

## Combining features
Since we have individual level data, we can combine the features arbitrarily using any stratification that we want.
This stratification can by any expression using the features in the `diseasystore`. Since we are working expressions,
we need to pass the expressions as "quosures" using `rlang::quos()`.

These expressions have very few limits. As long as they are translatable by `dbplyr` to SQL, they can be used.

Lets look at some examples:

### No stratification
```{r no_stratifications, eval = suggests_available}
# Get the number of positive tests by age group and sex
data1 <- ds$key_join_features(
  observable = "n_positive",
  stratification = NULL,
  start_date = ds$min_start_date,
  end_date = ds$max_end_date
)

print(data1)
```

```{r no_stratifications_plot, eval = suggests_available && rlang::is_installed("ggplot2")}
ggplot2::ggplot(data1, ggplot2::aes(x = date, y = n_positive)) +
  ggplot2::geom_line()
```

### Stratifying by a custom age group
```{r stratifications_positive_sex_age, eval = suggests_available}
# Get the number of positive tests by age group and sex
data2 <- ds$key_join_features(
  observable = "n_positive",
  stratification = rlang::quos(
    age_group = cut(
      age,
      breaks = c(0, 15, 30, Inf),
      labels = !!age_labels(c(0, 15, 30)),
      include.lowest = TRUE
    ),
    sex
  ),
  start_date = ds$min_start_date,
  end_date = ds$max_end_date
)

print(data2)
```

```{r stratifications_positive_sex_age_plot, eval = suggests_available && rlang::is_installed("ggplot2")}
ggplot2::ggplot(data2, ggplot2::aes(x = date, y = n_positive, color = age_group)) +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(~ sex)
```

### Stratifing by generation
```{r stratifications_admission_generation, eval = suggests_available}
# Get the number of admissions by generation
data3 <- ds$key_join_features(
  observable = "n_admission",
  stratification = rlang::quos(
    generation = dplyr::case_when(
      lubridate::year(birth) < 1946 ~ "Silent or older",
      lubridate::year(birth) < 1965 ~ "Boomer",
      lubridate::year(birth) < 1981 ~ "GenX",
      lubridate::year(birth) < 1997 ~ "Millenial",
      TRUE ~ "GenZ"
    )
  ),
  start_date = ds$min_start_date,
  end_date = ds$max_end_date
)

print(data3)
```

```{r stratifications_admission_generation_plot, eval = suggests_available && rlang::is_installed("ggplot2")}
ggplot2::ggplot(data3, ggplot2::aes(x = date, y = n_admission, color = generation)) +
  ggplot2::geom_line()
```

```{r cleanup, include = FALSE}
if (exists("ds")) rm(ds)
gc()
if (!rlang::is_installed("withr")) {
  options(opts)
}
```


[^1]: Notice that we don't use `{lubridate}` or syntax such as `birth_date + 365`.
As you may recall from the primer `vignette("extending-diseasystore")`, features are stored in data bases, not as R
objects. Therefore, the operations we perform must be translatable to SQL via `{dbplyr}` and date arithmetic are not
currently translatable consistently. `diseasystore` therefore provides some helper functions for the type of
date-related arithmetic that is common in feature computation. These are available for the data base back ends supported
by `diseasystore`.

[^2]: Notice that we split the `dplyr::mutate()` calls into several calls since SQL translation works differently from
"normal" `dplyr` behaviour. Normally in `dplyr`, the variable created earlier in one `dplyr::mutate()` call is available
immediately after, but this is not the case in SQL syntax. Therefore, we split the `dplyr::mutate()` calls which
generates nested sub-queries which makes the variables available in subsequent calls.

[^3]: `vignette("diseasystore")` explains the options system in more detail.

[^4]: The `DiseasystoreSimulist` is bundled with the package, but not exported. Therefore, we use the `:::` operator to
retrieve the class.
