---
title: "Extending diseasystore - simulist example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending diseasystore - simulist example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasystore)
```

```{r hidden_options, include = FALSE}
if (rlang::is_installed("withr")) {
  withr::local_options("tibble.print_min" = 5)
  withr::local_options("diseasystore.verbose" = FALSE)
} else {
  opts <- options("tibble.print_min" = 5, "diseasystore.verbose" = FALSE)
}

# We have a "hard" dependency for DuckDB to render parts of this vignette
suggests_available <- rlang::is_installed("duckdb")
```

# Introduction
If you haven't already, please consult the `vignette("extending-diseasystore")` which explains the concepts of the
`diseasystore`s in detail.

In this example, we go through how to create a `diseasystore` that implements individual level data.
For this purpose, we use a synthetic data set that we have generated using the `{simulist}` package.
This data set is a synthetic linelist from a disease outbreak.

```{r simulist_data}
simulist_data
```

Given such a data set, the aim is to create a `diseasystore` that can be used to retrieve features from this data set.


# Feature identification

To do this, we need to identify the "features" that we want to store in the `diseasystore`.
In the data, there are some demographic features such as "sex" and "age" and there are disease specific features such
as "date_onset", "date_admission", etc.

In this example we will implement the following features:

* Demographic features
  * `birth`: The birth date of the individuals
  * `age`: The (time-dependent) age of the individuals
  * `sex` - The sex of the individuals

* Disease features
  * `n_test` - The tests performed on the individuals
  * `n_positive` - The positive tests performed on the individuals
  * `n_admission` - The time of admission among the individuals
  * `n_hospital` - The time of hospitalisation among the individuals

Notice that the naming of the features differ by the prefix `n_` versus no prefix.
This dichotomy separates "stratifications" from "observables". Either is considered a "feature" in the diseasystore,
but the type of data (stratification vs. observable) is used when we join the data in `$key_join_features()`.
However, there are no differences in the mechanical implementation of the features --- only in the naming.


With the features identified, we begin by assigning a `FeatureHandler` to each feature, and design the map between
the feature names and the responsible `FeatureHandlers`:

```{r ds_map, eval = FALSE}
DiseasystoreSimulist <- R6::R6Class(
  classname = "DiseasystoreSimulist",
  inherit = DiseasystoreBase,

  ...

  private = list(
    .ds_map = list(
      "birth"       = "simulist_birth",
      "sex"         = "simulist_sex",
      "age"         = "simulist_age",
      "n_test"      = "simulist_test",
      "n_positive"  = "simulist_positive",
      "n_admission" = "simulist_admission",
      "n_hospital"  = "simulist_hospital"
    ),
    .label = "Simulist Synthetic Data",

  ...
)
```
Notice that we add the prefix `simulist_` to the feature names to avoid any potential conflicts with other features in
the database.

# Feature implementation
With the features identified, it is time to implement them in the `diseasystore`.

That is, we need to construct the `FeatureHandlers` for each feature.

## The `birth` `FeatureHandler`
We start by implementing the `birth` feature.

For this example, we will show first the code that implements the `FeatureHandler` and then describe the code in detail.
```{r feature_handler_birth, eval = FALSE}
  private = list(
    ...

    # The "birth" feature contains the birth dates of the individuals and is used later to compute
    # the age of the individuals at any given time.
    simulist_birth = FeatureHandler$new(
      compute = function(start_date, end_date, slice_ts, source_conn, ...) {

        out <- simulist_data |>
          dplyr::transmute(
            "key_pnr" = .data$id,
            "birth" = .data$birth,
            "valid_from" = .data$birth,
            "valid_until" = .data$date_death + lubridate::days(1)
          ) |>
          dplyr::filter({{ start_date }} < .data$valid_until, .data$valid_from <= {{ end_date }})

        return(out)
      },
      key_join = key_join_count
    ),

    ...
```

Notice that we create a new instance of `FeatureHandler` and assign it to the `simulist_birth` field in the private
list.

Lets focus on the `FeatureHandler` we create:
Here, we need to define how the feature is computed (that is, we implement the `compute()` function).

The signature of this function is fixed, and must take the arguments `start_date`, `end_date`, `slice_ts`,
`source_conn`, but can then take additional arguments in `...`.

The `birth` feature is a simple feature that we can extract directly from the data set.

In addition to the feature  values themselves (the values stored in the `birth` column) we need to include keying data
that tells `diseasystore` what other features this feature can be joined to.
Since we are working with individual level data, we can use a personal identification number, "pnr", as the key.
The simulist data already contains a unique ID for each individual, so we use this as the personal identification
number.

In addition, we need to specify the time period that the feature covers.
In this case, a birth date is assigned at birth, so we use this as the `valid_from` date.
For the expiration date of the information, we could either use `NA`, since the birth date is always valid, or
we could use the date of death, if the individual has died.
In this implementation we use the latter, since this allows for an easier implementation of the age of the individuals
later on.

Notice that `valid_until` is set as the date after death. The validity period of the features in `diseasystore` is
the interval $t \in \left[\textrm{valid_from}, \textrm{valid_until}\right)$.

After computing the feature, we filter to ensure that the data is returned only for the requested time period.

Finally, we also how the feature can be summarised. That is, we need to specify the type of `key_join` the feature
should use.

To illustrate how to determine the type of `key_join`, we imagine that we want to determine the number of people who
are born on each date. Given our data set, we could perform the following operations to summarize the data:

```{r key_join_birth}
simulist_data |>
  dplyr::count(lubridate::year(.data$birth))
```

That is, our feature has a record for each individual, and to summarise the feature, we need to `count` the individuals
within each group. This is different from semi-aggregated data where we would `sum` the values within each group.


## The `sex` `FeatureHandler`
We continue by implementing the `sex` feature.

Like the `birth` feature, this is a simple feature that we can mostly extract directly from the data set:

```{r feature_handler_sex, eval = FALSE}
  private = list(
    ...

    # The "sex" feature simply stores the sex from the simulist data
    simulist_sex = FeatureHandler$new(
      compute = function(start_date, end_date, slice_ts, source_conn, ds, ...) {

        out <- simulist_data |>
          dplyr::left_join(
            ds$get_feature("birth", start_date, end_date, slice_ts),
            by = c("id" = "key_pnr"),
            copy = TRUE
          ) |>
          dplyr::transmute(
            "key_pnr" = .data$id,
            "sex" = dplyr::if_else(.data$sex == "m", "Male", "Female"),
            .data$valid_from, .data$valid_until # Use values from birth feature
          ) |>
          dplyr::filter({{ start_date }} < .data$valid_until, .data$valid_from <= {{ end_date }})

        return(out)
      },
      key_join = key_join_count
    ),

    ...
```

Notice that we again create a new instance of `FeatureHandler` and assign it to the `simulist_sex` field in the
private list.

The computation of the feature is mostly straight forward, but we would like to use the validity period computed
in the `birth` feature (the same logic applies for the validity period of the "sex" information as the "birth"
information)

Therefore, we immediately see a difference in the function signature of the `compute()` function, where we no include
the `ds` argument which, like the others, are passed to the feature handler by `diseasystore` when we compute features.
This argument holds a reference to the `diseasystore` we are building, so we can use it to retrieve other features
(in this case, the `birth` feature).

Once we have the `birth` feature, we combine it with the `simulist_data` and extract a human-readable label for the
`sex`.

Finally, we filter the data like before to ensure that the data is returned only for the requested time period.


## The `age` `FeatureHandler`
The `age` feature is a bit more complex than the previous features, as it is a time-dependent feature.

Lets first look at the code:

```{r feature_handler_age, eval = FALSE}
  private = list(
    ...

    # The "age" feature computes the age of the individuals throughout the study period
    simulist_age = FeatureHandler$new(
      compute = function(start_date, end_date, slice_ts, source_conn, ds, ...) {

        # Using birth date, compute the age at the start of the study period
        age <- ds$get_feature("birth", start_date, end_date, slice_ts) |>
          dplyr::mutate(age_at_start = as.integer(!!age_on_date("birth", start_date, conn = ds %.% target_conn))) |>
          dplyr::compute()

        # Now, compute the next birth dates of the individual (as many as we need to cover the study period)
        # and compute the age of the individuals throughout the study period with their birth dates denoting the starts
        # and ends of the validity periods.
        out <- purrr::map(
          seq.int(from = 0, to = ceiling(lubridate::interval(start_date, end_date) / lubridate::years(1))),
          ~ age |>
            dplyr::mutate(
              "age" = .data$age_at_start + .x # The age for this iteration of the age computation loop
            ) |>
            dplyr::mutate( # We need to split the mutates to make the "age" column available for the next mutate
              "birthday" = !!add_years("birth", "age", conn = ds %.% target_conn) # Compute the birthday for the age
            ) |>
            dplyr::mutate( # Again, we need the split to make "birthday" available for the next mutate
              "next_birthday" = !!add_years("birthday", 1,  conn = ds %.% target_conn) # And when that age is not valid
            ) |>
            dplyr::filter( # Now we remove the birthdays that fall outside of the study period
              .data$birthday <= !!end_date,
              .data$birthday < .data$valid_until | is.na(.data$valid_until)
            ) |>
            dplyr::transmute( # We assign the birth dates as the validity periods
              "key_pnr" = .data$key_pnr,
              "age" = .data$age,
              "valid_from" = as.Date(.data$birthday),
              "valid_until" = pmin(.data$valid_until, as.Date(.data$next_birthday), na.rm = TRUE)
            )
        ) |>
          purrr::reduce(dplyr::union_all) # And we combine each age computation to a single age dataset

        return(out)
      },
      key_join = key_join_count
    ),

    ...
```

As with the `sex` feature, this feature also depends on the `birth` feature, so we include the `ds` argument in the
function signature of `compute()`.

Since the age is time-dependent, we need to compute the age of the individuals at the start of the study period (i.e.
the period requested by the user), and compute their age throughout the study period.

The age at the start of the study period is computed directly by from the birth date of the individuals using the
age helper function `age_on_date()`[^1].

Once the age at start is known, we determine the length of the study period (in years) and compute the birth days of
the individuals throughout the period[^2]. Each birth day has an associated age and a validity period which runs from the
birth day to the next birth day. Notice that we sometimes compute birth days that fall outside of the study period,
which we filter out before combining to a single data set with `dplyr::union_all`.

In all, we end with a data set containing the time-dependent age of the individuals throughout the study period.


## The `n_test` `FeatureHandler`

```{r feature_handler_n_test, eval = FALSE}
  private = list(
    ...

    # The "n_test" feature contains the tests taken by the individuals
    simulist_test = FeatureHandler$new(
      compute = function(start_date, end_date, slice_ts, source_conn, ...) {

        out <- simulist_data |>
          dplyr::filter(.data$case_type != "suspected") |>
          dplyr::transmute(
            "key_pnr" = .data$id,
            "valid_from" = .data$date_onset,
            "valid_until" = .data$valid_from + lubridate::days(1)
          ) |>
          dplyr::filter({{ start_date }} < .data$valid_until, .data$valid_from <= {{ end_date }})

        return(out)
      },
      key_join = key_join_count
    ),

    ...
```

## The `n_positive` `FeatureHandler`

```{r feature_handler_n_positive, eval = FALSE}
  private = list(
    ...

    # The "n_positive" feature contains the positive tests taken by the individuals
    simulist_positive = FeatureHandler$new(
      compute = function(start_date, end_date, slice_ts, source_conn, ...) {

        out <- simulist_data |>
          dplyr::filter(.data$case_type == "confirmed") |>
          dplyr::transmute(
            "key_pnr" = .data$id,
            "valid_from" = .data$date_onset,
            "valid_until" = .data$valid_from + lubridate::days(1)
          ) |>
          dplyr::filter({{ start_date }} < .data$valid_until, .data$valid_from <= {{ end_date }})


        return(out)
      },
      key_join = key_join_count
    ),

    ...
```

## The `n_hospital` `FeatureHandler`

```{r feature_handler_n_hospital, eval = FALSE}
  private = list(
    ...

    # The "n_hospital" feature contains the hospitalizations of the individuals
    simulist_hospital = FeatureHandler$new(
      compute = function(start_date, end_date, slice_ts, source_conn, ds, ...) {

        out <- simulist_data |>
          dplyr::filter(.data$case_type == "confirmed", !is.na(.data$date_admission)) |>
          dplyr::transmute(
            "key_pnr" = .data$id,
            "valid_from" = .data$date_admission,
            "valid_until" = .data$date_discharge + lubridate::days(1)
          ) |>
          dplyr::filter({{ start_date }} < .data$valid_until, .data$valid_from <= {{ end_date }})

        return(out)
      },
      key_join = key_join_count
    ),

    ...
```

## The `n_admission` `FeatureHandler`

```{r feature_handler_n_admission, eval = FALSE}
  private = list(
    ...

    # The "n_admission" feature contains the admissions of the individuals
    # We here use the "n_hospital" feature to compute the admissions since the admission is an entry for the
    # first date of hospitalisation
    simulist_admission = FeatureHandler$new(
      compute = function(start_date, end_date, slice_ts, source_conn, ds, ...) {

        out <- ds$get_feature("n_hospital", start_date, end_date, slice_ts) |>
          dplyr::mutate("valid_until" = .data$valid_from + 1L) |>
          dplyr::filter({{ start_date }} < .data$valid_until, .data$valid_from <= {{ end_date }})

        return(out)
      },
      key_join = key_join_count
    )

    ...
```


# The final product

The `diseasystores` require a database to store its features in.
These should be configured before use and can be stored in the packages options.
```{r configure_diseasystore, eval = FALSE}
# We define target_conn as a function that opens a DBIconnection to the DB
target_conn <- \() DBI::dbConnect(duckdb::duckdb())
options(
  "diseasystore.DiseasystoreSimulist.target_conn" = target_conn
)
```

```{r configure_diseasystore_hidden, include = FALSE, eval = suggests_available}
target_conn <- \() DBI::dbConnect(duckdb::duckdb())
if (rlang::is_installed("withr")) {
  withr::local_options("diseasystore.DiseasystoreSimulist.target_conn" = target_conn)
} else {
  opts <- c(opts, options("diseasystore.DiseasystoreSimulist.target_conn" = target_conn))
}
```


We can initialize the `diseasystore`
```{r initializing_diseasystore, eval = suggests_available}
ds <- diseasystore:::DiseasystoreSimulist$new()
```

Once configured such, we can use the feature store directly to get data.
```{r using_diseasystore_example_1, eval = suggests_available}
# We can see all the available features in the feature store
ds$available_features
```

```{r using_diseasystore_example_2, eval = suggests_available}
# And then retrieve a feature from the feature store
ds$get_feature(feature = "n_hospital",
               start_date = as.Date("2019-12-01"),
               end_date = as.Date("2020-01-13"))
```


```{r cleanup, include = FALSE}
if (exists("ds")) rm(ds)
gc()
if (!rlang::is_installed("withr")) {
  options(opts)
}
```


[^1]: Notice that we don't use `{lubridate}` or syntax such as `birth_date + 365`.
As you may recall from the primer `vignette("extending-diseasystore")`, features are stored in data bases, not as R
objects. Therefore, the operations we perform must be translatable to SQL via `{dbplyr}` and date arithmetic are not
currently translatable consistently. `diseasystore` therefore provides some helper functions for the type of
date-related arithmetic that is common in feature computation. These are available for the data base back ends supported
by `diseasystore`.

[^2]: Notice that we split the `dplyr::mutate()` calls into several calls since SQL translation works differently from
"normal" `dplyr` behaviour. Normally in `dplyr`, the variable created earlier in one `dplyr::mutate()` call is available
immediately after, but this is not the case in SQL syntax. Therefore, we split the `dplyr::mutate()` calls which
generates nested sub-queries which makes the variables available in subsequent calls.
