---
title: "diseasystore: Google COVID-19 data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diseasystore: Google COVID-19 data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasystore)
```

The Google COVID-19 data repository is a comprehensive open repository of COVID-19 data.
This vignette shows how to use this data with the `diseasy` and `diseasystore` packages.

First, it is a good idea to copy the relevant Google COVID-19 data files locally and store that location as an option for the package.
```{r download, eval=FALSE}
options(diseasystore.DiseasystoreGoogleCovid19.source_conn = "/local/file/path") # Should not end with "/"

# Define the Google files to download
google_files <- c("by-age.csv", "demographics.csv", "index.csv", "weather.csv")

# Download each file and compress them to reduce storage
#' \dontrun{
purrr::walk(google_files, ~ {
  url      <- paste0(getOption("diseasystore.DiseasystoreGoogleCovid19.remote_conn"), .)
  destfile <- file.path(getOption("diseasystore.DiseasystoreGoogleCovid19.source_conn"), .)
  download.file(url, destfile)
  R.utils::gzip(destfile)
})
#' }
```

The `diseasystores` require a data base to store its features in.
These should be configured before use and can be stored in the packages options.
NOTE: a global DB can also be configured using `options(diseasystore.target_conn = ...)` instead.

```{r init_diseasystore, eval=FALSE}

# We will store the target_conn as a function that opens a DBIconnection to the DB
target_conn <- \() dbConnect(RSQLite::SQLite(), "/local/file/path/diseasystore_google_covid_19.sqlite")
options(diseasystore.DiseasystoreGoogleCovid19.target_conn = target_conn)
```


Once the files are downloaded and the target DB is configured, we can run the `diseasystore` that uses the Google COVID-19 data.

```{r using_diseasystore, eval=FALSE}

# For ease of use, we can define a period of interest:
start_date <- as.Date("2020-01-01")
end_date   <- as.Date("2020-06-01")

fs <- DiseasystoreGoogleCovid19$new(start_date, end_date)
```
