---
title: "diseasystore: Google Health COVID-19 Open Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diseasystore: Google Health COVID-19 Open Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasystore)
```

The Google COVID-19 data repository is a comprehensive open repository of COVID-19 data.

This vignette shows how to use this data with the `diseasy` and `diseasystore` packages.

First, it is a good idea to copy the relevant Google COVID-19 data files locally and store that location as an option for the package.
```{r download_data, eval=FALSE}
# First we set the path we want to use as an option
options(diseasystore.DiseasystoreGoogleCovid19.source_conn = "/local/file/path")

# Ensure folder exists
if (!dir.exists(options() %.% diseasystore.DiseasystoreGoogleCovid19.source_conn)) {
  dir.create(options() %.% diseasystore.DiseasystoreGoogleCovid19.source_conn,
             recursive = TRUE, showWarnings = FALSE)
}

# Define the Google files to download
google_files <- c("by-age.csv", "demographics.csv", "index.csv", "weather.csv")

# Download each file and compress them to reduce storage
#' \dontrun{
purrr::walk(google_files, ~ {
  url <- paste0(getOption("diseasystore.DiseasystoreGoogleCovid19.remote_conn"), .)

  destfile <-
    file.path(getOption("diseasystore.DiseasystoreGoogleCovid19.source_conn"), .)

  if (!file.exists(paste0(destfile, ".gz"))) {
    download.file(url, destfile)
    R.utils::gzip(destfile)
  }
})
#' }
```

The `diseasystores` require a data base to store its features in.

These should be configured before use and can be stored in the packages options.

NOTE: a global DB can also be configured using `options(diseasystore.target_conn = ...)` instead.

```{r configure_diseasystore, eval=FALSE}

# We will store the target_conn as a function that opens a DBIconnection to the DB
target_conn <- \() DBI::dbConnect(RSQLite::SQLite(),
                                  "/local/file/path/diseasystore_google_covid_19.sqlite")

options(diseasystore.DiseasystoreGoogleCovid19.target_conn = target_conn)
```


Once the files are downloaded and the target DB is configured, we can run the `diseasystore` that uses the Google COVID-19 data.

```{r initializing_diseasystore, eval=FALSE}

# For ease of use, we can define a period of interest:
start_date <- as.Date("2020-01-01")
end_date   <- as.Date("2020-06-01")

fs <- DiseasystoreGoogleCovid19$new(start_date, end_date)
```

The feature store is designed to work with the `diseasy` package, which utilizes it to couple data as requested.
However, we can also use the feature store directly to get data.

```{r using_diseasystore, eval=FALSE}

# We can see all the available features in the feature store
print(fs$available_features)

# And then retrieve each feature from the feature store
#' \dontrun{
fs$get_feature("n_hospital")
#' }
```
